Evolution of ORM Systems: Formal Foundations

### 2013-01-26 doc/description.tex
merge classes

\paragraph{Merge of Classes} There occur situations when there is a need to merge two classes and the inlining transformation cannot be used because of name collisions among properties in such a case the $merge$ transformation can be used. It can be used in case structure of both classes (respectively tables) corresponds, the transformation does not change the structure, it only moves data. 
\begin{align*}
&	\mathbf{merge} : Label \times Label \times Label \times State \rightarrow State \\
&	\Phi(s, merge(l_1, l_2, l_n) = \inference{equals(c_1, c_2)}{application(s) \circ copyClassStructure(l_1, l_n)} \\\\
&	\Psi(s, merge(l_1, l_2, l_n) = \inference{\exists t_1, t_2 \in tables(database(s)) : \\ name(t_1) = l_1 \wedge name(t_2) = l_2 \\
\wedge ! \: referenced(t_1, d) \wedge ! \: referenced(t_2, d) }{database(s) \bullet mergeTables(l_1, l_2)}\\
& l_1, l_2, l_n \in Label, s \in State
\end{align*}

\section{Case Study}
The case study demonstrates how the transformation helps with data evolution in real example of software evolution. There are two classes in our software $Person$ and $LegalPerson$ both of them contains information about address (street, city and zip code). To improve the design of code the class Address has to be created which associated with both original classes and contains the data already stored in the software. The initial situation:
\begin{align*}
&	S = sw(a, db(t,r))
\end{align*}
\begin{align*}
a &= app(cl(Person, (pr(surname, APPSTRING, \varnothing), \\
& pr(street, APPSTRING, \varnothing), \\
& pr(city, APPSTRING, \varnothing)), pr(zip, APPSTRING, \varnothing)), \\
& cl(LegalPerson, (pr(businessName, APPSTRING, \varnothing), \\
& pr(street, APPSTRING, \varnothing), \\
& pr(city, APPSTRING, \varnothing)), pr(zip, APPSTRING, \varnothing))
\end{align*}
\begin{align*}
t &= db(tab(Person, pk("Id"), col(surname, APPSTRING, \varnothing, \varnothing), \\
& col(street, APPSTRING, \varnothing, \varnothing), col(city, APPSTRING, \varnothing, \varnothing) \\
& col(zip, APPSTRING, \varnothing, \varnothing), \varnothing), \\
& tab(LegalPerson, pk("Id"),  col(businessName, APPSTRING, \varnothing, \varnothing), \\
& col(street, APPSTRING, \varnothing, \varnothing), col(city, APPSTRING, \varnothing, \varnothing) \\
& col(zip, APPSTRING, \varnothing, \varnothing), \varnothing)) 
\end{align*}
\begin{align*}
r &= row("person", ["Id", 1], ["surname", "Griffin"], \\
& ["street", "S. Mary Evenue 22"], ["city", "NY"], ["zip", 10001]), \\
& row("person", ["Id", 2], ["surname", "Doe"], \\
& ["street", "Olive ave. 2225"], ["city", "LA"], ["zip", 90003]), \\
& row("legalPerson", ["Id", 1], ["businessName", "Tools and Machines ltd."], \\
& ["street", "Hill stret 22"], ["city", "NY"], ["zip", 10001]), \\
& row("legalPerson", ["Id", 2], ["businessName", "Doe"], \\
& ["street", "Pine ave. 2225"], ["city", "LA"], ["zip", 90003]))
\end{align*}
The first evolution step is to create the address class for $Person$ and $LegalPerson$, then the temporary class can be merged together and finally the temporary class has to be renamed to $Address$:
\begin{align*}
& state(S) \oplus Extract("Person", "Address_tmp1", "street", "city", "zip") \\
& \oplus Extract("LegalPerson", "Address_tmp2", "street", "city", "zip")  \\
& \oplus Merge("Address_tmp1", "Address_tmp2") \\
& \oplus Rename("Address_tmp2", "Address")
\end{align*}
The software state after second extraction is:

@@

\newpage
ExtactClass = addClass
\begin{align*}
a' &= app(cl(Person, (pr(surname, APPSTRING, \varnothing), \\
& pr(street, APPSTRING, \varnothing), \\
& pr(city, APPSTRING, \varnothing)), pr(zip, APPSTRING, \varnothing)), \\
& cl(LegalPerson, (pr(businessName, APPSTRING, \varnothing), \\
& pr(street, APPSTRING, \varnothing), \\
& pr(city, APPSTRING, \varnothing)), pr(zip, APPSTRING, \varnothing)),
& cl("Address_tmp1", \varnothing, \varnothing)
& )
\end{align*}
\begin{align*}
t &= db(tab(Person, pk("Id"), col(surname, APPSTRING, \varnothing, \varnothing), \\
& col(street, APPSTRING, \varnothing, \varnothing), col(city, APPSTRING, \varnothing, \varnothing) \\
& col(zip, APPSTRING, \varnothing, \varnothing), \varnothing), \\
& tab(LegalPerson, pk("Id"),  col(businessName, APPSTRING, \varnothing, \varnothing), \\
& col(street, APPSTRING, \varnothing, \varnothing), col(city, APPSTRING, \varnothing, \varnothing) \\
& col(zip, APPSTRING, \varnothing, \varnothing), \varnothing),
& tab("Address_tmp1", pk("Id"), \varnothing, \varnothing)) 
\end{align*}
\begin{align*}
r &= row("person", ["Id", 1], ["surname", "Griffin"], \\
& ["street", "S. Mary Evenue 22"], ["city", "NY"], ["zip", 10001]), \\
& row("person", ["Id", 2], ["surname", "Doe"], \\
& ["street", "Olive ave. 2225"], ["city", "LA"], ["zip", 90003]), \\
& row("legalPerson", ["Id", 1], ["businessName", "Tools and Machines ltd."], \\
& ["street", "Hill stret 22"], ["city", "NY"], ["zip", 10001]), \\
& row("legalPerson", ["Id", 2], ["businessName", "Doe"], \\
& ["street", "Pine ave. 2225"], ["city", "LA"], ["zip", 90003]))
\end{align*}

ExtactClass = copyProperty("street"
\begin{align*}
a' &= app(cl(Person, (pr(surname, APPSTRING, \varnothing), \\
& pr(street, APPSTRING, \varnothing), \\
& pr(city, APPSTRING, \varnothing)), pr(zip, APPSTRING, \varnothing)), \\
& cl(LegalPerson, (pr(businessName, APPSTRING, \varnothing), \\
& pr(street, APPSTRING, \varnothing), \\
& pr(city, APPSTRING, \varnothing)), pr(zip, APPSTRING, \varnothing)),
& cl("Address_tmp1", & pr(street, APPSTRING, \varnothing) , \varnothing)
& )
\end{align*}
\begin{align*}
t &= db(tab(Person, pk("Id"), col(surname, APPSTRING, \varnothing, \varnothing), \\
& col(street, APPSTRING, \varnothing, \varnothing), col(city, APPSTRING, \varnothing, \varnothing) \\
& col(zip, APPSTRING, \varnothing, \varnothing), \varnothing), \\
& tab(LegalPerson, pk("Id"),  col(businessName, APPSTRING, \varnothing, \varnothing), \\
& col(street, APPSTRING, \varnothing, \varnothing), col(city, APPSTRING, \varnothing, \varnothing) \\
& col(zip, APPSTRING, \varnothing, \varnothing), \varnothing),
& tab("Address_tmp1", pk("Id"), col(street, APPSTRING, \varnothing, \varnothing),\\
& \varnothing)) 
\end{align*}
\begin{align*}
r &= row("person", ["Id", 1], ["surname", "Griffin"], \\
& ["street", "S. Mary Evenue 22"], ["city", "NY"], ["zip", 10001]), \\
& row("person", ["Id", 2], ["surname", "Doe"], \\
& ["street", "Olive ave. 2225"], ["city", "LA"], ["zip", 90003]), \\
& row("legalPerson", ["Id", 1], ["businessName", "Tools and Machines ltd."], \\
& ["street", "Hill stret 22"], ["city", "NY"], ["zip", 10001]), \\
& row("legalPerson", ["Id", 2], ["businessName", "Doe"], \\
& ["street", "Pine ave. 2225"], ["city", "LA"], ["zip", 90003])\\
& row("address\_tmp1", ["Id", 1], ["street", "S. Mary Evenue 22"]) \\
\end{align*}

>>>>

\paragraph{Merge of Classes} There occur situations when there is a need to merge two classes and the inlining transformation cannot be used because of name collisions among properties in such a case the $merge$ transformation can be used. It can be used in case structure of both classes (respectively tables) is equals. The transformation creates new rows in the new table - the primary keys from origin instances are replaced by a new ones.
\begin{align*}
&	\mathbf{merge} : Label \times Label \times Label \times State \rightarrow State \\
&	\Phi(s, merge(l_1, l_2, l_n) = \inference{
		\exists c_1, c_2 \in classes(application(s)) \wedge 
		equals(c_1, c_2)}{\begin{gathered}
			application(s) \circ addClass(cl(ln, \varnothing, \varnothing)\\ \circ (\forall p \in properties(c_1) : addProperty(l_n, p)) \\ \circ (\forall a \in associations(c_1): addAssociation(l_n, a))
		\end{gathered}} \\\\
&	\Psi(s, merge(l_1, l_2, l_n) = \inference{\exists t_1, t_2 \in tables(database(s)) : \\ name(t_1) = l_1 \wedge name(t_2) = l_2 \\
\wedge ! \: referenced(t_1, d) \wedge ! \: referenced(t_2, d) }{\begin{gathered}
	 database(s) \bullet addTable(l_n, \mathbf{X}, \varnothing, \varnothing) \\ \bullet (\forall c \in columns(t_1) : addColumn(l_n, c)) \\ \bullet (\forall f \in foreignKeys(t_1) : addForeignKey(l_n, f)) \\ \bullet (\forall r \in instances(t_1, database(s)) \cup instances(t_2, database(s)) :  \\ insertRow(l_n, row(l_n, next(seq(key(l_n))), pairs(r)))) 
\end{gathered}}\\
& l_1, l_2, l_n \in Label, s \in State
\end{align*}

\section{Case Study}
The case study demonstrates how the transformation helps with data evolution in real example of software evolution. There are two classes in our software $Person$ and $LegalPerson$ both of them contains information about address (street, city and zip code). To improve the design of code the class Address has to be created which associated with both original classes and contains the data already stored in the software. The initial situation:
\begin{align*}
&	S = sw(a, db(t,r))
\end{align*}
\begin{align*}
a &= app(cl(Person, (pr(surname, APPSTRING, \varnothing), \\
& pr(street, APPSTRING, \varnothing), \\
& pr(city, APPSTRING, \varnothing)), pr(zip, APPSTRING, \varnothing)), \\
& cl(LegalPerson, (pr(businessName, APPSTRING, \varnothing), \\
& pr(street, APPSTRING, \varnothing), \\
& pr(city, APPSTRING, \varnothing)), pr(zip, APPSTRING, \varnothing))
\end{align*}
\begin{align*}
t &= db(tab(Person, pk("Id"), col(surname, APPSTRING, \varnothing, \varnothing), \\
& col(street, APPSTRING, \varnothing, \varnothing), col(city, APPSTRING, \varnothing, \varnothing) \\
& col(zip, APPSTRING, \varnothing, \varnothing), \varnothing), \\
& tab(LegalPerson, pk("Id"),  col(businessName, APPSTRING, \varnothing, \varnothing), \\
& col(street, APPSTRING, \varnothing, \varnothing), col(city, APPSTRING, \varnothing, \varnothing) \\
& col(zip, APPSTRING, \varnothing, \varnothing), \varnothing)) 
\end{align*}
\begin{align*}
r &= row("person", ["Id", 1], ["surname", "Griffin"], \\
& ["street", "S. Mary Evenue 22"], ["city", "NY"], ["zip", 10001]), \\
& row("person", ["Id", 2], ["surname", "Doe"], \\
& ["street", "Olive ave. 2225"], ["city", "LA"], ["zip", 90003]), \\
& row("legalPerson", ["Id", 1], ["businessName", "Tools and Machines ltd."], \\
& ["street", "Hill stret 22"], ["city", "NY"], ["zip", 10001]), \\
& row("legalPerson", ["Id", 2], ["businessName", "Doe"], \\
& ["street", "Pine ave. 2225"], ["city", "LA"], ["zip", 90003]))
\end{align*}
The first evolution step is to create the address class for $Person$ and $LegalPerson$, then the temporary class can be merged together and finally the temporary class has to be renamed to $Address$:
\begin{align*}
& state(S) \oplus Extract("Person", "Address_tmp1", "street", "city", "zip") \\
& \oplus AddAssociation("Person", assoc("address", "Address_tmp1", 1, 1),\\ 
& \{(x,y)| x \in instances("Person", database(S)), \\
& y \in instances("Address_tmp1", database(S)) : pkVal(x) = pkVal(y) \}) \\
& \oplus Extract("LegalPerson", "Address_tmp2", "street", "city", "zip")  \\
& \oplus AddAssociation("LegalPerson", assoc("address", "Address_tmp1", 1, 1),\\ 
& \{(x,y)| x \in instances("LegalPerson", database(S)), \\
& y \in instances("Address_tmp2", database(S)) : pkVal(x) = pkVal(y) \}) \\
& \oplus Merge("Address_tmp1", "Address_tmp2", "Address") \\
& \oplus RemoveTableAndData("Address_tmp1") \\
& \oplus RemoveTableAndData("Address_tmp2")
\end{align*}
The software state after adding the second association is:

@@

\newpage
ExtactClass = addClass
\begin{align*}
a' &= app(cl(Person, (pr(surname, APPSTRING, \varnothing), \\
& pr(street, APPSTRING, \varnothing), \\
& pr(city, APPSTRING, \varnothing)), pr(zip, APPSTRING, \varnothing)), \\
& cl(LegalPerson, (pr(businessName, APPSTRING, \varnothing), \\
& pr(street, APPSTRING, \varnothing), \\
& pr(city, APPSTRING, \varnothing)), pr(zip, APPSTRING, \varnothing)),
& cl("Address_tmp1", \varnothing, \varnothing)
& )
\end{align*}
\begin{align*}
t &= db(tab(Person, pk("Id"), col(surname, APPSTRING, \varnothing, \varnothing), \\
& col(street, APPSTRING, \varnothing, \varnothing), col(city, APPSTRING, \varnothing, \varnothing) \\
& col(zip, APPSTRING, \varnothing, \varnothing), \varnothing), \\
& tab(LegalPerson, pk("Id"),  col(businessName, APPSTRING, \varnothing, \varnothing), \\
& col(street, APPSTRING, \varnothing, \varnothing), col(city, APPSTRING, \varnothing, \varnothing) \\
& col(zip, APPSTRING, \varnothing, \varnothing), \varnothing), \\
& tab("Address_tmp1", pk("Id"), \varnothing, \varnothing)) 
\end{align*}
\begin{align*}
r &= row("person", ["Id", 1], ["surname", "Griffin"], \\
& ["street", "S. Mary Evenue 22"], ["city", "NY"], ["zip", 10001]), \\
& row("person", ["Id", 2], ["surname", "Doe"], \\
& ["street", "Olive ave. 2225"], ["city", "LA"], ["zip", 90003]), \\
& row("legalPerson", ["Id", 1], ["businessName", "Tools and Machines ltd."], \\
& ["street", "Hill stret 22"], ["city", "NY"], ["zip", 10001]), \\
& row("legalPerson", ["Id", 2], ["businessName", "Doe"], \\
& ["street", "Pine ave. 2225"], ["city", "LA"], ["zip", 90003]))
\end{align*}

ExtactClass = copyProperty("street"
\begin{align*}
a' &= app(cl(Person, (pr(surname, APPSTRING, \varnothing), \\
& pr(street, APPSTRING, \varnothing), \\
& pr(city, APPSTRING, \varnothing)), pr(zip, APPSTRING, \varnothing)), \\
& cl(LegalPerson, (pr(businessName, APPSTRING, \varnothing), \\
& pr(street, APPSTRING, \varnothing), \\
& pr(city, APPSTRING, \varnothing)), pr(zip, APPSTRING, \varnothing)), \\
& cl("Address_tmp1", pr(street, APPSTRING, \varnothing) , \varnothing)
& )
\end{align*}
\begin{align*}
t &= db(tab(Person, pk("Id"), col(surname, APPSTRING, \varnothing, \varnothing), \\
& col(street, APPSTRING, \varnothing, \varnothing), col(city, APPSTRING, \varnothing, \varnothing) \\
& col(zip, APPSTRING, \varnothing, \varnothing), \varnothing), \\
& tab(LegalPerson, pk("Id"),  col(businessName, APPSTRING, \varnothing, \varnothing), \\
& col(street, APPSTRING, \varnothing, \varnothing), col(city, APPSTRING, \varnothing, \varnothing) \\
& col(zip, APPSTRING, \varnothing, \varnothing), \varnothing), \\
& tab("Address_tmp1", pk("Id"), col(street, APPSTRING, \varnothing, \varnothing),\\
& \varnothing)) 
\end{align*}
\begin{align*}
r &= row("person", ["Id", 1], ["surname", "Griffin"], \\
& ["street", "S. Mary Evenue 22"], ["city", "NY"], ["zip", 10001]), \\
& row("person", ["Id", 2], ["surname", "Doe"], \\
& ["street", "Olive ave. 2225"], ["city", "LA"], ["zip", 90003]), \\
& row("legalPerson", ["Id", 1], ["businessName", "Tools and Machines ltd."], \\
& ["street", "Hill stret 22"], ["city", "NY"], ["zip", 10001]), \\
& row("legalPerson", ["Id", 2], ["businessName", "Doe"], \\
& ["street", "Pine ave. 2225"], ["city", "LA"], ["zip", 90003])\\
& row("address\_tmp1", ["Id", 1], ["street", "S. Mary Evenue 22"]) \\
\end{align*}
